---
title: "Tornado first regression"
author: "Helen Greatrex"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: yes
    number_sections: yes
    theme: lumen
    df_print: paged
  html_notebook:
    toc: true
    toc_float: yes
    number_sections: yes
    theme: lumen   
---

# Headings

## sub-headings

### sub-sub-heading (as long as there is a SPACE after the #)

we have space for text. Click insert (top right, then R)

```{r, message=FALSE, warning=FALSE}
# load library
# if a library doesn't exist, then to download, go to the packages menu (next to help) and click install
rm(list=ls())
library(sp)
library(sf)
library(raster)
library(tidycensus)
library(tidyverse)
library(plotly)
library(olsrr)
library(tmap)
library(RColorBrewer)
library("lubridate")
```

Write some more notes. Click preview to see a doc of the answer
Look at Rstudio markdown tutorial for more cool stuff

```{r}
# dot . means current folder - and you don't need a file extension
T_data <- st_read(dsn="./1950-2019-torn-initpoint",
                  layer="1950-2019-torn-initpoint")

```

To see the data, you can click on its name, or make a code chunk and type View(T_data)

We can access individual columns using the $ symbol

```{r}
# columnn names, by the way, in a code chunk, the # means don't run this line
names(T_data)
```

```{r}
summary(T_data)
```
```{r}
# print the 1st 6  values
head(T_data$date)
```

```{r}
head(T_data)
```



```{r}
# we need to find the format
# we look at the help file for strptime (this is not obvious)
#?strptime
```


```{r}
# make the date
T_data$date <- as.Date(T_data$date, format="%Y-%m-%d")
```

```{r}
summary(T_data)

```


```{r}
#  make a new column with date and time. paste always makes something a character
T_data$datetime <- paste(T_data$date,T_data$time)
```


```{r}
# and make this a timestamp
T_data$datetime <- as.POSIXlt(T_data$datetime )

```



```{r}
T_data$time <- paste("2002-07-15",T_data$time)
T_data$time <- as.POSIXlt(T_data$time )[[1]]

```



```{r}
# the tilda ~ means y depends on x... so fatalities (y axis) depend on time (x-axis)
# ignore time for now
plot(T_data$time,T_data$mag,cex=.2)
```

# remove missing data

mag < 0 has been set to NA, as directed by https://www.spc.noaa.gov/wcm/data/SPC_severe_database_description.pdf


```{r}
T_data$mag[T_data$mag < 0 ] <- NA
```



```{r}
# the tilda ~ means y depends on x... so fatalities (y axis) depend on time (x-axis)
# ignore time for now
plot(T_data$time,T_data$fat,cex=.2)
```

TO DO: Look at the exisitng papers on time of day, did they scale by per tornado?  or something else?





```{r}
# the tilda ~ means y depends on x... so fatalities (y axis) depend on time (x-axis)
# ignore time for now
plot(T_data$fat ~ T_data$mag,cex=.5)
```


this tutorial will be useful in honing code
https://hgreatrex.github.io/lab-7.html#tutorial-2-regression


```{r}
#x,y

p <-  T_data %>%                  
  ggplot( aes(mag,fat,col= time,label=om)) +
  geom_point() +
  theme_classic()+
  scale_color_gradient(low="blue", high="red")

ggplotly(p)
```


to do a regression

```{r}
firstmodel <- lm(fat ~ mag, data=T_data)
ols_regress(firstmodel)

firstmodel
```

fatalities = mag*0.73 - 0.9098


```{r}
plot(T_data$fat ~ T_data$mag)
abline(firstmodel)
```

```{r}
T_data$FirstModel.Residuals <- residuals(firstmodel)




# subset the illinois census data with the Chicago city limits
tm_shape(T_data, unit = "mi") + 
           tm_dots(col = "FirstModel.Residuals", style = "quantile",
                       palette = "-RdBu", border.alpha = 0, 
                       title = "First model residuals")#+
 #   tm_shape(Chicago.city) +
  #         tm_borders()+
    tm_layout(legend.outside = TRUE) 
```

